Nail: A practical interface generator for data formats

### 2014-11-14 paper/login/design.tex
Login

We will use DNS packets as a leading example, as defined in RFC 1035~\cite[\S4]{RFC:1035}.  Each DNS packet consists of a
header, a set of question records, and a set of answer records. Domain
names in both queries and answers are encoded as a sequence of labels,
terminated by a zero byte.  Labels are Pascal-style strings, consisting
of a length field followed by that many bytes comprising the label.
\label{s:design}
\subsection{Basic Data formats}
\input{fig-dnsfull}
Let's stepping through a simplified Nail grammar for non-compressed
packets, shown in  Figure~\ref{fig:dns-full}.  For this grammar, Nail produces the type declarations
shown in Figure~\ref{fig:dns-full-struct}, and the parser and
generator functions shown in Figure~\ref{fig:dns-full-api}. A more concise reference of the syntax
can be found in~\ref{fig:syntax}.
Nail grammars are reuseable between applications and we will use this grammar to implement
both a DNS server and client, which previously would have had two separate hand-written parsers,
leading to bugs such as  the Android Master Key.

@@

Just like C supports various constructs to build up types, such as structures, unions, pointers and
elemental types, Nail supports a integers as elemental parsers and a plethora of \emph{combinators}
to represent features of a file or protocol.

@@

The body of the \texttt{dnspacket} rule is a structure, which contains any number of fields enclosed
between  curly braces. Each field in the structure is parsed in sequence, and represented as a
structure to the programmer.  Contrary to other programming languages, Nail does not have a special
keyword for structs. We also reverse the usual structure-field syntax: \texttt{id uint1} is a field
called \texttt{id} with type \texttt{uint1}\footnote{This is a feature from Go, and makes nested
  types more readable and the parser for Nail itself easier to write.}.

\paragraph{Wrap combinator.} 
When implementing real protocols with Nail, we often found
structures that consist of many constants and only one named field. This pattern is
common in binary protocols which use fixed headers to denote the type of data
structure to be parsed.  In order to keep the internal representation cleaner,
we introduced the wrap combinator, which takes a sequence of parsers containing
exactly one non-constant parser. The external format is defined as though the wrap combinator were a
structure, but the data model does not introduce a structure with just one element, making the application-visible representation
(and thus application code) more concise.
Line 39 of the DNS grammar uses the wrap combinator to hide the terminating NUL-byte of a sequence
of labels.

\paragraph{Choices.}
If multiple structures can appear at a given position in a format, the programmer lists the options
along with a label for each in the \cc{choose} combinator. 
During parsing, Nail remembers the current input position and attempts each option in the order they
appear in the grammar. If an option fails, the parser backtracks to the initial position. If no
options succeed, the entire combinator fails. In the data model, choices are represented as tagged
unions.  The programmer has to be careful when options overlap, because if the programmer meant to
generate output for a choice, but the external representation is also valid for an earlier,
higher-priority option, the parser will interpret it as such. However, real data formats normally do
not have this overlap and we did not encounter it in the grammars we wrote.
An example is provided in Figure~\ref{fig:grammar-arith}.
\paragraph{Repetition.}

The \cc{many} combinator takes a parser and applies it repeatedly
until it fails, returning an array of the inner parser's results. In line 39 of the DNS grammar, a
sequence of labels is parsed by parsing as many labels as possible, that is, until an invalid length
field is encountered.
The \cc{sepBy} combinator
additionally takes a constant parser, which it applies in between parsing
two values, but not before parsing the first value or after parsing the
last.  This is useful for parsing an array of items delimited by a separator.
%For examle, \cc{many uint8} represents an array of 8-bit unsigned
\paragraph{Optional.}
Nail includes an \cc{optional} combinator, which attempts to recognize a value, but succeeds
without consuming input when it cannot recognize that value. Syntactically, \cc{optional} is
equivalent to a choice between the parser and an empty structure, but in the internal model it is
more concisely represented as a reference that is null when the parser fails.
For example, the grammar for  Ethernet headers uses \cc{optional vlan\_header} to parse the VLAN
header that appears only in Ethernet packets transmitted to a non-default VLAN\@.

\paragraph{References.}
Rules allow for recursive grammars, but some programming languages, such as C, do not support
recursive structures.  We introduce the
reference combinator \cc{*}  that does not change the syntax of the external format described, but
introduces a reference or pointer to the internal data type.
The reference combinator does not need to be used when another combinator, such as \cc{optional} or
\cc{many}, already introduces indirection in the data type. An example is shown in Figure~\ref{fig:grammar-arith}.

\begin{figure}
\smaller[0.5]
\begin{verbatim}
expr = choose {
  PAREN = <uint8='('; *expr; uint8=')'>
  PRODUCT = sepBy1 uint8='*' expr
  SUM = sepBy1 uint8='+' expr
  INTEGER = many1 uint8 | '0' .. '9'
}
\end{verbatim}
% \cc{paren = < uint8="("; optional *paren; uint8=")" >}.
\caption{Grammar for sums and products of integers.}
\label{fig:grammar-arith}
\end{figure}
% To support recursive structures, we include a reference combinator \cc{*}, which
%   is syntactically equivalent to directly including the parser, but in the internal model adds a
%   level of indirection to the value. This allows recursive structures to be parsed, f

\subsection{Redundant Data}
\label{s:dependent}
Data formats often contain values that are determined by other values or the layout of information,
such as checksums, duplicated information, or offset and  length fields. Exposing such values risks
inconsistencies that could trick the program into unsafe behaviour. Therefore, we represent such values using \emph{dependent fields} and handle them transparently during
parsing and generation without exposing them to the application.

In DNS packets, the packet header contains count fields (\cc{qc},
\cc{ac}, \cc{ns}, and \cc{ar}), which contain the number of questions and answers that follow the
header, which we represent by dependent fields (line 12-15).

Dependent fields are defined within a structure like normal fields, but their name starts with an \cc{@} symbol.
A dependent field is in scope and can be referred to by the definition of all subsequent fields in
the same structure. Dependent fields can be passed to rule invocations as parameters.

Dependent fields are handled like other fields when parsing input, but only stored temporarily
instead of in the data type. Their value can be referenced by other parsers until it goes out of scope.
When generating output, Nail visits a dependent field twice. First, while generating the other fields of a
structure, the generator reserves space for the dependent field in the output. Once
the dependent field goes out of scope, the generator  writes the
dependent field's value to this space.

@@

Handling label compression in existing tools, such as Bison or Hammer,
would be awkward at best, because some ad-hoc trick would have to
be used to re-position the parser's input stream.  Keeping track of the
position of all recognized labels would not be enough, although still awkward, as the offset field
may refer to any byte within the packet.
For this reason, the sample DNS grammar in Hammer does not
support compression.

@@

>>>>

We will use DNS packets as a leading example, as defined in RFC 1035~\cite[\S4]{RFC:1035}.  Each DNS packet consists of a
header, a set of question records, and a set of answer records. Domain
names in both queries and answers are encoded as a sequence of labels,
terminated by a zero byte.  Labels are Pascal-style strings, consisting
of a length field followed by that many bytes comprising the label.
\label{s:design}
\subsection{Basic Data formats}

\input{fig-dnsfull}
Let's stepping through a simplified Nail grammar for non-compressed
packets, shown in  Figure~\ref{fig:dns-full}.  For this grammar, Nail produces the type declarations
shown in Figure~\ref{fig:dns-full-struct}, and the parser and
generator functions shown in Figure~\ref{fig:dns-full-api}.
Nail grammars are reuseable between applications and we will use this grammar to implement
both a DNS server and client, which previously would have had two separate hand-written parsers,
leading to bugs such as  the Android Master Key.

@@

Just like C supports various constructs to build up types, such as structures, unions, pointers and
elemental types, Nail supports a integers as elemental parsers and a plethora of \emph{combinators}
to represent features of a file or protocol. We will present the features we used in implementing
DNS. A more complete reference can be found in~\ref{fig:syntax}, with detailed rationale in~\cite{bangert:nail-osdi14}.

@@

The body of the \texttt{dnspacket} rule is a structure, which contains any number of fields enclosed
between  curly braces. Each field in the structure is parsed in sequence, and represented as a
structure to the programmer.  Contrary to other programming languages, Nail does not have a special
keyword for structs. We also reverse the usual structure-field syntax: \texttt{id uint1} is a field
called \texttt{id} with type \texttt{uint1}\footnote{This is a feature from Go, and makes nested
  types more readable and the parser for Nail itself easier to write.}. Often, Nail grammars have
structures with just one non-constant field, for example when parsing a fixed header. To prevent
this from cluttering the data type, we support an alternative form of the structure combinator,
using angle brackets, that has just one unnamed, non-constant field, which is represented directly
in the data type.

% To support recursive structures, we include a reference combinator \cc{*}, which
%   is syntactically equivalent to directly including the parser, but in the internal model adds a
%   level of indirection to the value. This allows recursive structures to be parsed, f

\input{fig-syntax}

\subsection{Redundant Data}
\label{s:dependent}
Data formats often contain values that are determined by other values or the layout of information,
such as checksums, duplicated information, or offset and  length fields. Exposing such values risks
inconsistencies that could trick the program into unsafe behaviour. Therefore, we represent such values using \emph{dependent fields} and handle them transparently during
parsing and generation without exposing them to the application. 
Dependent fields are handled like other fields when parsing input, but only stored temporarily
instead of in the data type. Their value can be referenced by other parsers until it goes out of scope.
When generating output, Nail inserts the correct value.

In DNS packets, the packet header contains count fields (\cc{qc},
\cc{ac}, \cc{ns}, and \cc{ar}), which contain the number of questions and answers that follow the
header, which we represent by dependent fields (line 12-15).
Dependent fields are defined within a structure like normal fields, but their name starts with an \cc{@} symbol.
A dependent field is in scope and can be referred to by the definition of all subsequent fields in
the same structure. Dependent fields can be passed to rule invocations as parameters.

@@

\begin{figure}
\smaller[0.5]
\input{code/dns-xform}
\caption{Signatures of stream transform functions for handling DNS label compression.}
\label{fig:dns-xform}
\end{figure}
Handling label compression in existing tools, such as Bison or Hammer,
would be awkward at best, because some ad-hoc trick would have to
be used to re-position the parser's input stream.  Keeping track of the
position of all recognized labels would not be enough, although still awkward, as the offset field
may refer to any byte within the packet.
For this reason, the sample DNS grammar in Hammer does not
support compression.

@@

\begin{figure}[h]
\smaller[0.5]
\input{code/offset-xform}
\caption{Pseudocode for two functions that implement the offset transform.}
\label{fig:xform-sig}
\end{figure}

### 2014-11-14 paper/login/intro.tex
Login

However, as memory-safe languages and exploit mitigation  tricks are becoming more prevalent,
attackers are moving to a new class of attack - parser differentials. Many applications use
hand-written input processing code, which is often mixed with the rest of the application, e.g. by
passing a pointer to the raw input through the application. This (anti-)pattern makes it impossible
to figure out whether two implementations of the same format or protocol are identical, and input
handling code can't be easily reused between applications. As a result, different applications often
disagree in the corner cases of a protocol, which can have fatal security consequences. For example,
Android has two parsers for ZIP archives involved in securely installing applications. First, a Java
program checks the signatures of files contained within an app archive and then another tool
extracts them to the filesystem. Because the two ZIP parsers disagree in multiple places, attackers
can modify a valid file so that the verifier will see the original contents, but attacker-controlled
files will be extracted to the filesystem, bypassing Android's code signing. Similar issues showed
up on iOS \cite{evasion} and SSL \cite{layercake}.

@@

This approach is typical in compiler design and in other applications handling text-based inputs, but not
common for binary inputs. The Hammer framework~\cite{hammer} and data description languages such as
PADS~\cite{pads} have been developing generated parsers for binary protocols.

@@

Furthermore, some parser generators are cumbersome to use when parsing binary data for several reasons. First, parser generators typically parse inputs into an abstract syntax tree that corresponds to the grammar. Application developers must then write explicit {\em semantic actions} to construct a data structure the rest of the program can process. Therefore, programmers must describe the format up to three times -- in the grammar, the data structure and the semantic actions. Second, many applications also need to produce outputs in the same formats they parse - say when saving files or responding to network requests. However, most parser generators only address parsing inputs, so  the programmer must manually construct outputs. Some parser generators, such as 
Boost.Spirit~\cite{boost-spirit} allow generating output, but require the programmer to write another set of semantic actions.

We address these challenges with Nail, a new parser generator  with several features geared towards
binary formats.
First, Nail grammars describe not only a format, but also a data type to represent it within the program.
This removes the semantic actions and type declarations that programmers have to write with existing
parser generators. This forces the programmer
to clearly separate syntactic validation and semantic processing.

Second, Nail does not just
parse input into an internal representation, but also generates output from the internal representation,
without requiring the programmer to write additional code.

>>>>

However, as memory-safe languages and exploit mitigation  tricks are becoming more prevalent,
attackers are moving to a new class of attack - parser differentials. Many applications use
hand-written input processing code, which is often mixed with the rest of the application, e.g. by
passing a pointer to the raw input through the application. This (anti-)pattern makes it impossible
to figure out whether two implementations of the same format or protocol are identical, and input
handling code can't be easily reused between applications. As a result, different applications often
disagree in the corner cases of a protocol, which can have fatal security consequences. For example,
Android has two parsers for ZIP archives involved in securely installing applications. First, a Java
program checks the signatures of files contained within an app archive and then another tool
extracts them to the filesystem. Because the two ZIP parsers disagree in multiple places, attackers
can modify a valid file so that the verifier will see the original contents, but attacker-controlled
files will be extracted to the filesystem, bypassing Android's code signing. Similar issues showed
up on iOS \cite{geohot-evasion} and SSL \cite{DBLP:conf/fc/KaminskyPS10}.

@@

This approach is typical in compiler design and in other applications handling text-based inputs, but not
common for binary inputs. The Hammer framework~\cite{hammer-parser} and data description languages such as
PADS~\cite{Fisher:2005:PDL:1064978.1065046} have been developing generated parsers for binary protocols.

@@

Furthermore, some parser generators are cumbersome to use when parsing binary data for several
reasons. First, many parser generators don't directly data structures, but call semantic actions
that the programmer writes to construct a data structure. Therefore, programmers must
describe the format up to three times -- in the grammar, the data structure and the semantic
actions. Second, most parser generators only address parsing inputs, so programmers have to manually
construct outputs. Some parser generators, such as 
Boost.Spirit~\cite{boost-spirit} allow generating output, but require the programmer to write another set of semantic actions.

We address these challenges with Nail, a new parser generator for binary formats.
First, Nail grammars describe not only a format, but also a data type to represent it within the program.
This removes the semantic actions and type declarations that programmers have to write with existing
parser generators. It also forces the programmer to clearly separate syntactic validation and semantic processing.
Second, Nail does not just
parse input into an internal representation, but also generates output from the internal representation,
without requiring the programmer to write additional code.
